Appendix 3: Estimating species interactions

### 2015-11-23 Harris_2015/manuscript-materials/response-to-editor.md

> Dear Dr. de Valpine,

> [[stuff]]

> Sincerely,

> David J. Harris

@@

>For this reason, I now include three different simulation regimes with different properties for the 20-species landscapes: 1. Simulating directly from a Markov network, 2. Simulating from a Markov network with spatial heterogeneity in the $\alpha$ parameters (sometimes called a conditional random field), and 3. Simulating from a process that included abundances and per-capita effects rather than species-level interactions between binary species.  In order to avoid the problems that you highlighted with the previous submission, I simulated the abundances for the third set of landscapes using Gibbs sampling rather than a population dynamic model. To keep the abundances finite without adding arbitrary constraints to the new simulations, I removed mutualistic interactions from the third set of landscapes.

> The overall result is that the reader can now see how the different models perform when A) the species' "true" joint distribution is fully described by univariate and bivariate potentials (e.g. a Markov network), when B) the joint distribution also depends on external environmental factors, and when C) the strength of the interaction between two species depends on (unobserved) variation in abundance.

Note that often violations of assumptions do not create bias in parameter estimates (although sometimes they do) but rather incorrect assessments of uncertainty (inaccurate confidence intervals and Type I error rates in hypothesis tests). You have referred in multiple places to having applied a bootstrap as well as calculating Wald intervals, but it appears to me you have reported almost no results about validity of confidence intervals or hypothesis tests from those (there are a few results for the 3-species case). It would be important to do so.

* I now report approximate Type I error rates and the coverage of the model's 95% confidence intervals.  The 95% confidence intervals have about 98% coverage when the data was generated from an unmodified Markov network, and about 86% coverage in the presence of environmental heterogeneity (lines 201-206 and Appendix 5). These coverage values aren't perfect, but they are reasonable, and they provide far better inferences than ecologists' current methods (lines 206-213; Figure 4C).

@@

It appears to me that you did not understand my point that you are presenting results that are integrated over parameter distributions, which appears still to be the case. What I meant is the following. Often one would want to see how a statistical method performs as a function of specific parameter values. For example, one might vary the value of an interaction coefficient (beta) from negative (competition) to zero to positive (mutualism) and run a set of simulations for each value of beta. This would allow evaluation of statistical power and accuracy of confidence intervals and Type I error rate as well as RMSE, all as a function of true parameter values. It would also allow assessment of whether the method works better for some parameter scenarios than others, which is not uncommon. Instead you are drawing parameters from distributions and then presenting performance assessments in aggregate from those results, such as the average R-squared over all cases of a given network size. Generally, and clearly in your case, such metrics can be written as expectations (integrations) over the distribution of parameters you sampled from. You are right there is nothing Bayesian about it, but I pointed out a Bayesian may feel justified in doing it because many Bayesian results take the same format. It is a question of presenting aggregated versus disaggregated results. Unfortunately, presenting only aggregated results leaves the reader unable to dig deeper into the disaggregated results. The analogy to sampling from an archipelago is lovely but irrelevant. It appears to me that since you have all the simulations at hand, and since you can place more detailed results in the supplement, you should do so. I leave it to you to decide how to do that.

* Showing the results' dependence on the value of beta is a great idea, and I have added several analyses along these lines in Figure 4C and Appendix 5.  In particular, I now show how CI coverage, RMSE, and the probability that the model will predict the wrong sign for an interaction vary as a function of beta (in addition to global estimates of R-squared and Type I error rates).  If you think that readers would benefit from even more disaggregation, I could include a data file containing the seven models' estimates for all the species pairs across all 450 simulated landscapes as a sixth Appendix.

@@

> Beyond these steps, I've considered several other options for further exploration of the parameter space, but none of them would be very useful given the large number of parameters in these models (20 alphas and 190 betas). Systematically varying one of the beta coefficients while holding the other 209 parameters constant at arbitrary values would not convincingly yield generalizable insights, as different lines through the 210-dimensional space could produce different results.

Some other minor notes

I don't think you have stated that the betas are symmetric. It appears that beta_ij = beta_ji and hence each could be interpreted as half of the interaction strength. E.g. on line 80, it would be e^4 if I take the preceding model equation as correct and include both beta_ij and beta_ji. Please fix it up if that is not right.

* I've made this issue much clearer in the revision: see lines 68-78.  It should now be unambiguous that the sum in the likelihood should only include each pair once, and that each pair must share a single value for $\beta_{ij}$.

@@

 Both times that I have seen the title anew, the phrase "estimating species interactions from observational data" made me think it would be about time-series method, since that is a common kind of "observational data" from which people try to estimate interactions. I recommend adding a descriptive term like "species composition data" or "observational occurrence data" or "presence/absence data." You decide. This could enter the first Abstract sentence as well.

* This is another good point.  I've changed the title to focus on co-occurrence data.  Now that I've included inferential statistics in the paper and evaluated their frequentist properties, I've also changed the word "estimating" back to "inferring".

Line 44 is not stated well. You could replace "begin with the assumption" by "test the hypothesis" or insert "that the hypothesis of interest is that [all pairwaise interactions...]" or some such change.

* I have changed the wording to emphasize hypothesis testing rather than assumptions (lines 40-42).

@@

Line 69: The phrase "how groups of species can co-occur" seems imprecise or incorrect. Maybe "to determine species occurence models"?

* Lines 65-66 now describe "how... conditional relationships can determine species composition" rather than referring to "groups".

Lines 86-87 are not a good lead-in to the point of the rest of this paragraph. It sounds like you are about to talk about species interactions, again, but really the point of the paragraph is that normalizing the probabilities involves a difficult summation.

* The discussion of computational difficulties has been removed to save space, and the lines you noted are now at the end of a paragraph rather than the beginning of one (lines 82-85).

@@

Lines 72 & 101. I was willing to overlook the fact that you didn't define the vector y explicitly in terms of its elements on line 72 because perhaps that is really obvious. However on line 101 you then state maximum likelihood as if it would be done from a single y vector (and a single likelihood term) rather than a matrix of observations. So now I think you need more explicit notation to clarify the arrangements of your variables.

* The revised manuscript now clarifies that each $\vec{y}$ vector represents the different species that could occur at a single site and that the full co-occurrence matrix is made of "a set of independent $\vec{y}$ vectors indicating which species are present at each site on the landscape" (lines 89-90). This should clarify the relationship between the $\vec{y}$ vectors and the full matrix as well as the relationship between the single-site likelihood of Equation 1 and the landscape-level likelihood (i.e. independence among sites implies that the full likelihood equals the sum of the site-level likelihoods).

@@

Line 130-132. This statement is unclear and unconvincing. Are you saying that the remoteness of the upper range of the prior justifies that the prior has no influence? That would not be correct since it is an issue of prior weightings in the range of heavy posterior weight. The simplest way to show a prior has little influence is to try a couple of different ones.

* I no longer suggest that the prior has no influence. I also make no claims (e.g. regarding unbiasedness) that would be invalidated by a non-flat prior, so the strength of the prior should not affect the validity of any of the results presented. Your initial concern about the prior in my first submission was that the logistic distribution with scale 1 was too similar to the exponential distribution with rate 1 used to simulate the "true" distribution.  Now that the prior is more than twice as wide as the "true" distribution, this is no longer an issue.

> With regard to your suggestion that I try different priors, I have now used three different ones and gotten similar results: a flat prior (for the three-species landscapes where the MLE was always finite), and two different logistic priors (for the larger communities).

@@

Lines 136-137. Bootstrapping is a method to estimate uncertainty of an estimation procedure. It is unclear what you mean by using it to validate if a procedure gives "stable" estimates.

* The bootstrapping results have now been removed and replaced with the approximate confidence intervals.

@@

Figures in general: would benefit from titles on each subfigure.

* Subfigures now have titles

It's too bad we are in this unusual situation that your effort to respond to a reviewer made the manuscript worse. I will let you make another revision to get back on course. What I need you to do are: (1) remove the community dynamics model, put back in extensive results from the model of interest, and take a new approach to probing the model more thoroughly by including some scenarios with clearly defined violations of assumptions and/or data generated from one of the alternative statistical models. You may decide what to put in the online material since I recognize you are trying to stick to the 20-page report limit.

* As described above, I have removed the community dynamics model and now compare all the models' performance on simulated landscapes where all the assumptions are met, and in two cases where different assumptions are violated in clear ways.

@@

2 Include meaningful results about accuracy of uncertainties. These should be at your fingertips since you can calculate confidence intervals for every simulation.

* As described above, I now estimate the coverage of the confidence intervals, and (Appendix 5) show how coverage varies as a function of the "true" parameter value.

@@

3 Provide some disaggregated results.

* As described above, I now show how several performance metrics vary as a function of $\beta$ and have offered to include the raw results for all 7 methods across the 450 simulations.

@@

4 Show how the regression steps work in about a page or two of supplement text, not requiring code reading.

* I wasn't entirely sure whether you were referring to the logistic regression models or to the process of fitting the Markov network, but I believe that I've clarified the discussion of both issues sufficiently that supplemental material is no longer required for either:

### 2015-11-25 Harris_2015/manuscript-materials/response-to-editor.md

> Dear Dr. de Valpine,

> Thank you for your insightful and thorough comments.  I have addressed each of them below.

@@

>For this reason, I now include three different simulation regimes with different properties for the 20-species landscapes: 1. Simulating directly from a Markov network, 2. Simulating from a Markov network with spatial heterogeneity in the $\alpha$ parameters (sometimes called a conditional random field), and 3. Simulating from a process that included abundances and per-capita effects rather than species-level interactions between binary species.  In order to avoid the problems that you highlighted with the previous submission, I simulated the abundances for the third set of landscapes using Gibbs sampling rather than a population dynamic model. To keep the abundances finite without adding arbitrary constraints to the new simulations, I removed mutualistic interactions from the third set of landscapes.

> The overall result is that the reader can now see how the different models perform when A) the species' "true" joint distribution is fully described by univariate and bivariate potentials (e.g. a Markov network), when B) the joint distribution also depends on external environmental factors, and when C) the strength of the interaction between two species depends on (unobserved) variation in abundance.

Note that often violations of assumptions do not create bias in parameter estimates (although sometimes they do) but rather incorrect assessments of uncertainty (inaccurate confidence intervals and Type I error rates in hypothesis tests). You have referred in multiple places to having applied a bootstrap as well as calculating Wald intervals, but it appears to me you have reported almost no results about validity of confidence intervals or hypothesis tests from those (there are a few results for the 3-species case). It would be important to do so.

* I now report approximate Type I error rates and the coverage of the model's 95% confidence intervals.  The 95% confidence intervals have about 98% coverage when the data was generated from an unmodified Markov network, and about 86% coverage in the presence of environmental heterogeneity (lines 201-206 and Appendix 4). These coverage values aren't perfect, but they are reasonable, and they provide far better inferences than ecologists' current methods (lines 206-213; Figure 4C).

@@

It appears to me that you did not understand my point that you are presenting results that are integrated over parameter distributions, which appears still to be the case. What I meant is the following. Often one would want to see how a statistical method performs as a function of specific parameter values. For example, one might vary the value of an interaction coefficient (beta) from negative (competition) to zero to positive (mutualism) and run a set of simulations for each value of beta. This would allow evaluation of statistical power and accuracy of confidence intervals and Type I error rate as well as RMSE, all as a function of true parameter values. It would also allow assessment of whether the method works better for some parameter scenarios than others, which is not uncommon. Instead you are drawing parameters from distributions and then presenting performance assessments in aggregate from those results, such as the average R-squared over all cases of a given network size. Generally, and clearly in your case, such metrics can be written as expectations (integrations) over the distribution of parameters you sampled from. You are right there is nothing Bayesian about it, but I pointed out a Bayesian may feel justified in doing it because many Bayesian results take the same format. It is a question of presenting aggregated versus disaggregated results. Unfortunately, presenting only aggregated results leaves the reader unable to dig deeper into the disaggregated results. The analogy to sampling from an archipelago is lovely but irrelevant. It appears to me that since you have all the simulations at hand, and since you can place more detailed results in the supplement, you should do so. I leave it to you to decide how to do that.

* Showing the results' dependence on the value of beta is a great idea, and I have added several analyses along these lines in Figure 4C and Appendix 4.  In particular, I now show how CI coverage, RMSE, and the probability that the model will predict the wrong sign for an interaction vary as a function of beta (in addition to global estimates of R-squared and Type I error rates).  If you think that readers would benefit from even more disaggregation, I could include a data file containing the seven models' estimates for all the species pairs across all 450 simulated landscapes as a fifth Appendix.

@@

It's too bad we are in this unusual situation that your effort to respond to a reviewer made the manuscript worse. I will let you make another revision to get back on course. What I need you to do are: (1) remove the community dynamics model, put back in extensive results from the model of interest, and take a new approach to probing the model more thoroughly by including some scenarios with clearly defined violations of assumptions and/or data generated from one of the alternative statistical models. You may decide what to put in the online material since I recognize you are trying to stick to the 20-page report limit.

* As described above, I have removed the community dynamics model and now compare all the models' performance on simulated landscapes where all the assumptions are met, and in two cases where different assumptions are violated in clear ways.

2 Include meaningful results about accuracy of uncertainties. These should be at your fingertips since you can calculate confidence intervals for every simulation.

* As described above, I now estimate the coverage of the confidence intervals, and (Appendix 4) show how coverage varies as a function of the "true" parameter value.

### 2015-11-25 Harris_2015/manuscript-materials/draft.Rmd
as resubmitted

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 4).
First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
models, as described above. For the analyses with 20 species, a weak regularizer
(equivalent to a logistic prior with scale 2) ensured that the estimates were
always finite.

@@

\noindent
\textbf{Twenty species.}
In general, each model's performance was highest for large landscapes with
simple assembly rules and no environmental heterogeneity (Figure 3).
Despite some variability across contexts, the rank ordering across methods was
very consistent.  In particular, the four methods that controlled for indirect
effects (the Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did not.
The Markov network consistently performed best of all. As anticipated by
@lee_learning_2012, generalized linear models closely
approximated the Markov network estimates (Figure 4A), especially when the data
sets were very large (Figure 3). As reviewed in @gotelli_empirical_2009,
however, most analyses in this field of ecology involve fewer than 50 sites,
and the gap between the Markov network and GLMs is larger in this context. As
shown in Appendix 5, the standard errors associated with the estimates in Figure
3 are small (less than 0.01), so the differences among methods should not be
attributed to sampling error.

@@

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, for which the true interaction rates are easiest to
interpret).  The Markov network's approximate Type I error rate (defined here
as the probability that 0 fell outside the 95% confidence interval for a pair of
species where $|\beta_{ij}|<0.1$) depended on the simulation type: 0.02 for
simulations that matched the model's assumptions, versus 0.14 for simulations
that included environmental heterogeneity (see Appendix 5 for confidence
interval coverage across a range of $\beta_{ij}$ values). The null model's
Type I error rates were 0.30 and 0.50 for the constant and heterogeneous
landscapes, respectively---far higher than the nominal 0.05 rate. Figure
4C shows, across a range of true interaction strengths, the probability that
the null model or the Markov network will predict the wrong sign of the
interaction with 95% confidence.  The null model makes such errors more than 5 times
as often as the Markov network, even though it only reject the null hypothesis
twice as often overall (Appendix 5).  The Markov network's errors were also
more concentrated around 0, so it never misclassified strong interactions like
the null model did.

>>>>

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
models, as described above. For the analyses with 20 species, a weak regularizer
(equivalent to a logistic prior with scale 2) ensured that the estimates were
always finite.

@@

\noindent
\textbf{Twenty species.}
In general, each model's performance was highest for large landscapes with
simple assembly rules and no environmental heterogeneity (Figure 3).
Despite some variability across contexts, the rank ordering across methods was
very consistent.  In particular, the four methods that controlled for indirect
effects (the Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did not.
The Markov network consistently performed best of all. As anticipated by
@lee_learning_2012, generalized linear models closely
approximated the Markov network estimates (Figure 4A), especially when the data
sets were very large (Figure 3). As reviewed in @gotelli_empirical_2009,
however, most analyses in this field of ecology involve fewer than 50 sites,
and the gap between the Markov network and GLMs is larger in this context. As
shown in Appendix 4, the standard errors associated with the estimates in Figure
3 are small (less than 0.01), so the differences among methods should not be
attributed to sampling error.

@@

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, for which the true interaction rates are easiest to
interpret).  The Markov network's approximate Type I error rate (defined here
as the probability that 0 fell outside the 95% confidence interval for a pair of
species where $|\beta_{ij}|<0.1$) depended on the simulation type: 0.02 for
simulations that matched the model's assumptions, versus 0.14 for simulations
that included environmental heterogeneity (see Appendix 4 for confidence
interval coverage across a range of $\beta_{ij}$ values). The null model's
Type I error rates were 0.30 and 0.50 for the constant and heterogeneous
landscapes, respectively---far higher than the nominal 0.05 rate. Figure
4C shows, across a range of true interaction strengths, the probability that
the null model or the Markov network will predict the wrong sign of the
interaction with 95% confidence.  The null model makes such errors more than 5 times
as often as the Markov network, even though it only reject the null hypothesis
twice as often overall (Appendix 4).  The Markov network's errors were also
more concentrated around 0, so it never misclassified strong interactions like
the null model did.

### 2015-11-25 Harris_2015/manuscript-materials/draft.tex
as resubmitted

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a set of independent \(\vec{y}\) vectors indicating which species are
present at each site on the landscape). When the number of species is
reasonably small (less than about 30), one can find exact maximum
likelihood estimates for all of the \(\alpha\) and \(\beta\)
coefficients given a presence-absence matrix by numerically optimizing
\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces (Murphy 2012),
ensuring that this procedure will converge on the global maximum. I used
the rosalia package (Harris 2015a) for the R programming language (R
Core Team 2015) to calculate \(p(\vec{y}; \alpha, \beta)\) and its
gradients; the package passes these functions to the ``BFGS'' method in
R's general-purpose optimizer, which then estimates the Markov network
parameters.

@@

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 4).
First, I used the rosalia package (Harris 2015a) to fit Markov network
models, as described above. For the analyses with 20 species, I added a
weak regularizer (equivalent to a logistic prior with scale 2) to ensure
that the model estimates were always finite.

@@

The third alternative, generalized linear models (GLMs), also provide a
computationally efficient approximation to the Markov network (Lee and
Hastie 2012). Following Faisal et al. (2010), I fit regularized logistic
regression models (Gelman et al. 2008) for each species, using the other
species on the landscape as predictors. This produced two interaction
estimates for each species pair; one for the effect of species \(i\) on
species \(j\) and one for the reverse. These two coefficients are not
identifiable from the data, however (Schmidt and Murphy 2012), so I used
their average as an overall measure of their relationship.

@@

\noindent
\textbf{Twenty species.} In general, each model's performance was
highest for large landscapes with simple assembly rules and no
environmental heterogeneity (Figure 3). Despite some variability across
contexts, the rank ordering across methods was very consistent. In
particular, the four methods that controlled for indirect effects (the
Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did
not. The Markov network consistently performed best of all. As
anticipated by Lee and Hastie (2012), generalized linear models closely
approximated the Markov network estimates, especially when the data sets
were very large {[}{[}Figure{]}{]}. As reviewed in Gotelli and Ulrich
(2009), however, most analyses in this field of ecology involve fewer
than 50 sites, so model performance on the smaller landscapes (where the
gap was larger) is very important. As shown in Appendix 5, the standard
errors associated with the estimates in Figure 3 are small (less than
0.01), so the differences among methods should not be attributed to
sampling error.

@@

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, for which the true interaction rates are
easiest to interpret). The Markov network's approximate Type I error
rate (defined here as the probability that 0 fell outside the 95\%
confidence interval for a pair of species where \(|\beta_{ij}|<0.1\))
depended on the simulation type: 0.02 for simulations that matched the
model's assumptions, versus 0.14 for simulations that included
environmental heterogeneity (see Appendix 5 for confidence interval
coverage across a range of \(\beta_{ij}\) values). The null model's Type
I error rates were {[}{[}XX{]}{]} and {[}{[}XX{]}{]} for the constant
and heterogeneous landscapes, respectively---far higher than the nominal
0.05 rate. Figure 4C shows, across a range of true interaction
strengths, the probability that the null model or the Markov network
will predict the wrong sign of the interaction with 95\% confidence. The
null model makes such errors {[}{[}XX{]}{]} times as often as the Markov
network, even though it only reject the null hypothesis {[}{[}XX{]}{]}
as often overall (Appendix 5). The Markov network's errors were also
more concentrated around 0, so it never misclassified strong
interactions like the null model did.

@@

\noindent
The results presented above show that Markov networks can reliably
recover species' pairwise interactions from species composition data,
even for cases where environmental heterogeneity and indirect
interactions cause ecologists' typical null modeling approaches to
reliably fail. Partial covariances and generalized linear models can
both provide computationally efficient approximations, but with somewhat
lower accuracy, especially for typically-sized data sets with small
numbers of sites (Gotelli and Ulrich 2009). The difference in accuracy
may be larger for real data sets than for the simulated landscapes in
Figure 3, however; linear approximations to the Markov network make
larger errors when the interaction matrix is structured (e.g.~due to
guilds or trophic levels; Loh and Wainwright 2013). Similarly, the
separate generalized linear models for each species can severely overfit
in some cases (Lee and Hastie 2012). The full Markov network should thus
be preferred to the approximations when it is computationally tractable.

Compositional data only contains enough degrees of freedom to estimate
one interaction per species pair (Schmidt and Murphy 2012), so none of
these methods can identify the exact nature of the pairwise interactions
(e.g.~which species in a positively-associated pair is facilitating the
other). To estimate asymmetric interactions, such as commensalism or
predation, ecologists could use time series, behavioral observations,
manipulative experiments, or natural history. These other sources of
information could also be used to augment the likelihood function with
an informative prior distribution, reducing ecologists' error and
uncertainty relative to Figure 3's results.

@@

\setlength{\parindent}{0cm}

@@

\textbf{Figure 3:} Proportion of variance in interaction coefficients
explained by each method versus number of sampled locations across the
three simulation types. For the null model (Pairs), these values include
the removal of two outliers with \(|Z|>1000\) {[}{[}Appendix{]}{]}.

@@

\hyperdef{}{ref-xiaoux5fstrongux5f2015}{\label{ref-xiaoux5fstrongux5f2015}}
Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
\end{flushleft}
\end{spacing}

>>>>

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a set of independent \(\vec{y}\) vectors indicating which species are
present at each site on the landscape). When the number of species is
reasonably small (less than about 30), one can find exact maximum
likelihood estimates for all of the \(\alpha\) and \(\beta\)
coefficients given a presence-absence matrix by numerically optimizing
\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces (Murphy 2012),
ensuring that this procedure will converge on the global maximum. I used
the rosalia package (Harris 2015a) for the R programming language (R
Core Team 2015) to calculate \(p(\vec{y}; \alpha, \beta)\) and its
gradients (Murphy 2012); the package passes these functions to the
``BFGS'' method in R's general-purpose optimizer, which then estimates
the Markov network parameters.

@@

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package (Harris 2015a) to fit Markov network
models, as described above. For the analyses with 20 species, a weak
regularizer (equivalent to a logistic prior with scale 2) ensured that
the estimates were always finite.

@@

The third alternative, generalized linear models (GLMs), also provide a
computationally efficient approximation to the Markov network (Lee and
Hastie 2012). Following Faisal et al. (2010), I fit regularized logistic
regression models (Gelman et al. 2008) for each species, using the other
species on the landscape as predictors. This produced two interaction
estimates for each species pair; one for the effect of species \(i\) on
species \(j\) and one for the reverse. These two coefficients are not
identifiable from the data, however (Schmidt and Murphy 2012), so I used
their average as an overall measure of the pairwise relationship.

@@

\noindent
\textbf{Twenty species.} In general, each model's performance was
highest for large landscapes with simple assembly rules and no
environmental heterogeneity (Figure 3). Despite some variability across
contexts, the rank ordering across methods was very consistent. In
particular, the four methods that controlled for indirect effects (the
Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did
not. The Markov network consistently performed best of all. As
anticipated by Lee and Hastie (2012), generalized linear models closely
approximated the Markov network estimates (Figure 4A), especially when
the data sets were very large (Figure 3). As reviewed in Gotelli and
Ulrich (2009), however, most analyses in this field of ecology involve
fewer than 50 sites, and the gap between the Markov network and GLMs is
larger in this context. As shown in Appendix 4, the standard errors
associated with the estimates in Figure 3 are small (less than 0.01), so
the differences among methods should not be attributed to sampling
error.

@@

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, for which the true interaction rates are
easiest to interpret). The Markov network's approximate Type I error
rate (defined here as the probability that 0 fell outside the 95\%
confidence interval for a pair of species where \(|\beta_{ij}|<0.1\))
depended on the simulation type: 0.02 for simulations that matched the
model's assumptions, versus 0.14 for simulations that included
environmental heterogeneity (see Appendix 4 for confidence interval
coverage across a range of \(\beta_{ij}\) values). The null model's Type
I error rates were 0.30 and 0.50 for the constant and heterogeneous
landscapes, respectively---far higher than the nominal 0.05 rate. Figure
4C shows, across a range of true interaction strengths, the probability
that the null model or the Markov network will predict the wrong sign of
the interaction with 95\% confidence. The null model makes such errors
more than 5 times as often as the Markov network, even though it only
reject the null hypothesis twice as often overall (Appendix 4). The
Markov network's errors were also more concentrated around 0, so it
never misclassified strong interactions like the null model did.

@@

\noindent
The results presented above show that Markov networks can reliably
recover species' pairwise interactions from species composition data,
even for cases where environmental heterogeneity and indirect
interactions cause ecologists' typical null modeling approaches to
reliably fail. Partial covariances and generalized linear models can
both provide computationally efficient approximations, but with somewhat
lower accuracy (especially for typically-sized data sets with small
numbers of sites; Gotelli and Ulrich 2009). The difference in accuracy
may be larger for real data sets than for the simulated landscapes in
Figure 3, however; linear approximations to the Markov network make
larger errors when the interaction matrix is structured (e.g.~due to
guilds or trophic levels; Loh and Wainwright 2013). Similarly, the
separate generalized linear models for each species can severely overfit
in some cases (Lee and Hastie 2012). The full Markov network should thus
be preferred to the approximations when it is computationally tractable.

Compositional data only contains enough degrees of freedom to estimate
one interaction per species pair (Schmidt and Murphy 2012), so none of
these methods can identify the exact nature of the pairwise interactions
(e.g.~which species in a positively-associated pair is facilitating the
other). To estimate asymmetric interactions, such as commensalism or
predation, ecologists could use time series, behavioral observations,
manipulative experiments, or natural history. These other sources of
information could also be used to augment the likelihood function with a
more informative prior distribution, reducing ecologists' error and
uncertainty relative to Figure 3's results.

@@

\setlength{\parindent}{0cm}
\setlength{\parskip}{0pt}

@@

@@

\hyperdef{}{ref-xiaoux5fstrongux5f2015}{\label{ref-xiaoux5fstrongux5f2015}}
Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.

\noindent \textbf{Figures:}

\textbf{Figure 3:} Proportion of variance in interaction coefficients
explained by each method versus number of sampled locations across the
three simulation types. For the null model (Pairs), two outliers with
\(|Z|>1000\) were manually adjusted to \(|Z|=50\) to mitigate their
detrimental influence on \(R^2\) (Appendix 5).

\end{flushleft}
\end{spacing}

### 2015-11-25 Harris_2015/recovery.Rmd
fix naming
as resubmitted

---
output: 
  pdf_document:
    keep_tex: true
title: "Appendix 3: Estimating species interactions"
subtitle: \"Estimating species interactions from observational data with Markov networks\"
author: David J. Harris
fontsize: 11pt
---

@@

@@

mc.cores = 2

@@

Load in the results from the `pairs` program, run outside of R with
the following options:

@@

  splitname = strsplit(identifier, "/|-|\\.")[[1]]
  n_spp = as.integer(splitname[[1]]) # simulated species, NOT present species
  n_sites = as.integer(splitname[[2]])
  rep_name = splitname[[3]]

@@

```{r, cache = FALSE, eval = TRUE}

>>>>

---
output: 
  pdf_document:
    keep_tex: true
title: "Appendix 3: Estimating species interactions"
subtitle: "Inferring species interactions from co-occurrence data with Markov networks"
author: David J. Harris
fontsize: 11pt
---

@@

Note that the `pairs` program was run separately (outside of R) with the following options:

@@

mc.cores = 8

@@

@@

  splitname = strsplit(identifier, "/|-|\\.")[[1]]
  n_sites = as.integer(splitname[[1]])
  rep_name = splitname[[2]]

@@

```{r, cache = FALSE, eval = FALSE}
